
# (APPENDIX) Appendix {-}
```{r, echo=FALSE}
system("source ~/.bash_profile")
```

# Useful functions
## Gamma Function

We want to generalize the factorial function which was 
$$n! = n(n-1)(n-2)\dots(2)(1)$$
for $n \in \{0, 1, 2, \dots\}$ and we define $0!=1$ for notational convenience because that is what we need in many formulas.  Notice that $1! = 1$, $2!=2$ and $3! = 3 * 2!$.

The problem is that this is only defined for the whole numbers (natural numbers and zero).  We want to define a function that allows us to expand this function to the positive real numbers.

Consider the funtion:
$$\Gamma(x) = \int_0^\infty z^{x-1} e^{-z} \,dz$$

First we show that $\Gamma(x) = (x-1)\Gamma(x-1)$ utilizing integration by parts.

$$\begin{aligned}  
u  = z^{x-1}      &\;\;\;  & dv = e^{-z}\,dz \\
du = (x-1)z^{x-2}\,dx &\;\;\;  & v  = -e^{-z} 
\end{aligned}$$

and therefore

$$\begin{aligned}
\Gamma(x) &= \int_0^\infty z^{x-1} e^{-z} \,dz \\
          &= \int_0^\infty u \;dv \\
          &= uv\vert_0^\infty - \int_0^\infty v\, du \\
          &= -z^{x-1}e^{-z}\vert_{z=0}^\infty + \int_0^\infty (x-1)z^{x-2}e^{-z} \, dz\\
          &= (x-1) \int_0^\infty z^{x-2}e^{-z}\,dz\\
          &= (x-1) \Gamma(x-1)
  \end{aligned}$$

Next we notice that $\Gamma(1) = 1 = 0!$ via the following integration
$$\Gamma(1) = \int_0^\infty z^{1-1}e^{-z}\,dz = \int_0^\infty e^{-z} \,dz = -e^{-z} \vert_0^\infty = 1$$

Using these bits, we can see that 
$$\Gamma(2) = 1\cdot \Gamma(1) = 1 = 1!$$
$$\Gamma(3) = 2 \Gamma(2) = 2 = 2!$$
$$\Gamma(4) = 3 \Gamma(3) = 3 \cdot 2! = 3!$$
and for positive integers $n$,
$$\Gamma(n) = (n-1)!$$
which can be re-arranged to see that
$$n! = n \Gamma(n)$$

We conclude by noting that 
$$\Gamma\left( \frac{1}{2} \right) = \sqrt \pi$$ 
_(This result can be obtained by looking at the square of the integral and then doing a trig substitution.)_

A graph of this on the postive real numbers is given below: 

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='jpeg'}
library(dplyr)
library(ggplot2)
data <- data.frame(x=seq(0,6, by=.005)) %>%
  filter(x > 0) %>%
  mutate( y = gamma(x) )
ggplot(data, aes(x=x, y=y)) +
  geom_line() +
  coord_cartesian(xlim=c(0,5), ylim=c(0,12)) +
  scale_y_continuous(name=expression(Gamma(x)))
```

Finally we note that there is a reasonable approximation (known as Stirling's approximation) to the function for large values of $x$ as
$$\Gamma(x+1) \approx \sqrt{ 2 \pi x } \left( \frac{x}{e} \right)^x$$